---
title: "session1"
author: "Tim Riffe"
date: "2026-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

When I write here, this is just like writing in Word, or whatever.

Ctrl + Alt + i to make an R chunk. Cmnd + option + i on mac.

What we write inside this R chunk is executed behind the scenes, and it might produce output, and it might update objects that we have in memory for this session
```{r}
library(tidyverse)
# install.packages("tidyverse")
B <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_B2014.csv")
D <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_D2014.csv")
P <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_P2014.csv")

```

# timeout to install latex

This is a cheap solution for being able to build your document to pdf output, incase LaTeX isn't installed already on your machine:
```{r}
# install.packages("tinytex")
```
Note, I comment out the `install.packages()` line because otherwise it would reinstall each time I compile this, which is annoying.

# task 1: calculate exposures

 `|>` is a basic R _pipe_, and it works very similarly to `%>%` which might be the default pipe in.
```{r}
P_long <- 
  P |> 
  select(-open_interval) |> 
  pivot_longer(c(female, male, total),
               names_to = "sex",
               values_to = "pop")
```

First visualization of the exercise: put year 2014 and 2015 next to each other (`pivot_wider()`) and then take the mean `mutate()`. Pro: this might be how you'd do it in a spreadsheet? Easy to wrap your mind around.

```{r}
P_long |> 
  pivot_wider(names_from = year,
              values_from = pop) |> 
  mutate(exposure = (`2014` + `2015`) / 2)
```

Cons: what if we have a longer time series? This gets very wide and arduous to type out! We need another strategy:

General approach (use `lead()`)

```{r}
x <- 0:10
lead(x)

E <- 
  P_long |> 
  arrange(sex, age, year) |> 
  group_by(sex, age) |> 
  mutate(pop2 = lead(pop)) |> 
  filter(!is.na(pop2)) |> 
  mutate(exposure = (pop + pop2) / 2)
```















