---
title: "session1"
author: "Tim Riffe"
date: "2026-07-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

When I write here, this is just like writing in Word, or whatever.

Ctrl + Alt + i to make an R chunk. Cmnd + option + i on mac.

What we write inside this R chunk is executed behind the scenes, and it might produce output, and it might update objects that we have in memory for this session
```{r}
library(tidyverse)
# install.packages("tidyverse")
B <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_B2014.csv")
D <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_D2014.csv")
P <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ES_P2014.csv")

```

# timeout to install latex

This is a cheap solution for being able to build your document to pdf output, incase LaTeX isn't installed already on your machine:
```{r}
# install.packages("tinytex")
```
Note, I comment out the `install.packages()` line because otherwise it would reinstall each time I compile this, which is annoying.

# task 1: calculate exposures

 `|>` is a basic R _pipe_, and it works very similarly to `%>%` which might be the default pipe in.
```{r}
P_long <- 
  P |> 
  select(-open_interval) |> 
  pivot_longer(c(female, male, total),
               names_to = "sex",
               values_to = "pop")
```

First visualization of the exercise: put year 2014 and 2015 next to each other (`pivot_wider()`) and then take the mean `mutate()`. Pro: this might be how you'd do it in a spreadsheet? Easy to wrap your mind around.

```{r}
P_long |> 
  pivot_wider(names_from = year,
              values_from = pop) |> 
  mutate(exposure = (`2014` + `2015`) / 2)
```

Cons: what if we have a longer time series? This gets very wide and arduous to type out! We need another strategy:

General approach (use `lead()`)

```{r}
x <- 0:10
lead(x)

E <- 
  P_long |> 
  arrange(sex, age, year) |> 
  # group_by(sex, age) |> 
  mutate(pop2 = lead(pop),
         .by = c(sex, age)) |>  
  filter(!is.na(pop2)) |> 
  mutate(exposure = (pop + pop2) / 2) # same as ungroup()
```

Usually, classically, I would do `group_by(bla, bla) |> do thing |> ungroup()`. But the `ungroup()` part can be handled with `.groups = "drop"` inside a summarize call and then you don't need `ungroup()` OR, and this is even more parsimonious, you can just use groups temporarily by using the `.by` arg of `mutate()` or `summarize()`. `.by` is now part of the main tidyverse corpus, and is no longer experimental, so it's now the easiest choice unless of course you want to recycle groups over sequential operations.

# task 2: calculate fertility rates

```{r}
E |> 
  filter(sex == "total") |> 
  left_join(B, by = c("age","year")) |> 
  mutate(total = if_else(is.na(total), 0, total)) |> 
  summarize(B = sum(total),
            E = sum(exposure)) |> 
  mutate(CBR = 1000 * B / E)
```

# exercise: calculate age-specific fertility rates
These have females in the denominator, match to age, and note you want the age range given in the births file (`B`). This will require `filter()`, `*_join()` and `mutate()`. Keep it all together in a single pipeline :-). 5 min. Someone paste the solution here: <https://tinyurl.com/6ec2zfyy>
```{r}
asfr <- E |> 
  filter(sex == "female") |> 
  right_join(
    B |> select(-year, -open_interval),
    by = join_by(age)
  ) |>
  mutate(
    asfr = total / exposure
  )
```

visualize them:

```{r}
asfr |> 
  ggplot() +
  aes(x = age, y = asfr) +
  geom_step(linewidth = 1.05)

```

```{r}
asfr |> 
  summarize(tfr = sum(asfr))
```

# Exercise: calculate age-specific mortality rates
Join exposure and death data (by age and sex), calculate rates. Note, we still haven't reshaped the death data! Then make a log10 plot of the mortality rates :-)
 
 
```{r}
mort <- 
  D |> 
  select(-open_interval) |> 
  pivot_longer(female:total,
               names_to = "sex",
               values_to = "deaths") |> 
  left_join(E, by = join_by(year, age, sex)) |> 
  select(!starts_with("pop")) |> 
  mutate(mx = deaths / exposure) 
```
 
We can calculate the CDR with this too:

```{r}
cdr <-
mort |> 
  summarize(CDR = 1000 * sum(deaths) / sum(exposure),
            .by = c(sex))
```

# visualize age-specific rates:

```{r}
mort |> 
  filter(mx>0) |> 
  ggplot(mapping = aes(x = age, y = mx, color = sex)) +
  geom_line() +
  scale_y_log10() +
  # fixed!
  geom_hline(mapping = aes(yintercept = CDR/1000, color = sex), data = cdr)
```

# NEW: how do the age-specific sex weights change?

```{r}
mort |> 
  select(sex,age,exposure) |> 
  pivot_wider(names_from = sex, values_from = exposure) |> 
  mutate(p_male = male / total) |> 
  ggplot(aes(x = age)) +
  geom_ribbon(aes(ymin= rep(0,111),ymax=p_male), fill = "#11A1FF") +
  geom_ribbon(aes(ymin = p_male, ymax = rep(1,111)), fill = "#BB00A1") +
  labs(y = "fraction") +
  annotate("text",x=13,y=.9,label="Females",color = "white",size=8) +
  annotate("text",x=13,y=.1,label="Males",color = "white",size=8)
```

# Exercise
 
Calculate TFR and MAB by year for these data:

```{r, message = FALSE}
library(tidyverse)
library(janitor)
B <- read_csv("https://github.com/timriffe/BSSD2026Module2/raw/master/data/ES_B.csv.gz",
              show_col_types = FALSE) |> 
  clean_names()
P <- read_csv("https://github.com/timriffe/BSSD2026Module2/raw/master/data/ES_P.csv.gz",
              show_col_types = FALSE)|> 
  clean_names()
```

You'll need to calculate exposures, merge w births, then calculate rates. Recall TFR is:

$$ TFR = \sum_x F_x$$
where $F_x$ are age-specific fertility rates (ASFR). Mean age at childbearing (MAB) is a weighted average of... age. The thing being weighted is age itself, and the weights are either rates or births (results will vary a bit). The formula is something like this:

$$ MAB = \frac{\sum (age + 0.5)\cdot w}{\sum w}$$
where $w$ is either `asfr` or `births`. You can make TFR and MAB in the same `summarize()` call...

Then please plot them. I will also cover solutions from the handout for day 1.









