---
title: "day 2 session script"
author: "Tim Riffe"
date: "2026-07-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Today: lifetables and functions

let's see the 
`cumsum()`: makes a cumulative sum
`cumprod()`: makes a cumulative product
`rev()`: flips a vector
```{r}
x <- 1:10

cumsum(x)
cumprod(x)
rev(x)
rev(x) |> cumsum()
```

Anatomy of a function

```{r}
my_fun <- function(a, b){
  # the body of the function, where
  # things happen using the arguments
  
  # This is arbitrary code, just to show you
  # can use the args and other functions at-will
  d <- cumsum(a)
  bb <- cumprod(b)
  
  result <- sum(bb * d)
  # the only things allowed out of here are
  # what you explicitly return at the end.
  return(result)
}
my_fun(a = runif(10), b = sort(rnorm(10)))
```
# read in the data

<https://tinyurl.com/6ec2zfyy>
```{r}
library(tidyverse)
morty <- read_csv("https://github.com/timriffe/BSSD2026Module2/raw/refs/heads/master/data/ES_mort.csv.gz",
                  show_col_types = FALSE)
```

# Step 1

To calculate condition probabilities, $q_x$, we start with $m_x$ and an instrumental lifetable column denoted $a_x$.

This is valid for single-age intervals
$$ q_x = \frac{m_x}{1 + (1 - a_x)\cdot m_x}$$

$a_x$ has a meaning: it's the average years lived in the age interval by those that die in the interval.

```{r}

mx_to_ax <- function(mx, age, sex = "f"){
  ax <- case_when(
    age == 0 & sex == "m" & mx < 0.023 ~ 0.14929 - 1.99545 * mx,
    age == 0 & sex == "m" & mx >= 0.023 & mx < 0.08307 ~ 0.02832 + 3.26021 * mx,
    age == 0 & sex == "m" & mx >= 0.08307 ~ 0.29915,
    age == 0 & sex == "f" & mx < 0.01724 ~ 0.14903 - 2.05527 * mx,
    age == 0 & sex == "f" & mx >= 0.01724 & mx < 0.06891 ~ 0.04667 + 3.88089 * mx,
    age == 0 & sex == "f" & mx >= 0.06891 ~ 0.31411,
    age == max(age) ~ 1/mx,
    TRUE ~ 0.5
  )
  return(ax)
}


```























