---
title: "day 2 session script"
author: "Tim Riffe"
date: "2026-07-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Today: lifetables and functions

let's see the 
`cumsum()`: makes a cumulative sum
`cumprod()`: makes a cumulative product
`rev()`: flips a vector
```{r}
x <- 1:10

cumsum(x)
cumprod(x)
rev(x)
rev(x) |> cumsum()
```

Anatomy of a function

```{r}
my_fun <- function(a, b){
  # the body of the function, where
  # things happen using the arguments
  
  # This is arbitrary code, just to show you
  # can use the args and other functions at-will
  d <- cumsum(a)
  bb <- cumprod(b)
  
  result <- sum(bb * d)
  # the only things allowed out of here are
  # what you explicitly return at the end.
  return(result)
}
my_fun(a = runif(10), b = sort(rnorm(10)))
```
# read in the data

<https://tinyurl.com/6ec2zfyy>
```{r}
library(tidyverse)
morty <- read_csv("https://github.com/timriffe/BSSD2026Module2/raw/refs/heads/master/data/ES_mort.csv.gz",
                  show_col_types = FALSE)
```

# Step 1

To calculate condition probabilities, $q_x$, we start with $m_x$ and an instrumental lifetable column denoted $a_x$.

This is valid for single-age intervals
$$ q_x = \frac{m_x}{1 + (1 - a_x)\cdot m_x}$$

$a_x$ has a meaning: it's the average years lived in the age interval by those that die in the interval.

```{r}
mx_to_ax <- function(mx, age, sex = "f"){
    sex <- rep(sex[1], 111)
    ax = case_when(
    age == 0 & sex == "m" & mx < 0.023 ~ 0.14929 - 1.99545 * mx,
    age == 0 & sex == "m" & mx >= 0.023 & mx < 0.08307 ~ 0.02832 + 3.26021 * mx,
    age == 0 & sex == "m" & mx >= 0.08307 ~ 0.29915,
    age == 0 & sex == "f" & mx < 0.01724 ~ 0.14903 - 2.05527 * mx,
    age == 0 & sex == "f" & mx >= 0.01724 & mx < 0.06891 ~ 0.04667 + 3.88089 * mx,
    age == 0 & sex == "f" & mx >= 0.06891 ~ 0.31411,
    age == max(age) ~ 1/mx,
    TRUE ~ 0.5
  )
  return(ax)
}
```

# Exercise: write a function that does this:

$$ q_x = \frac{m_x}{1 + (1 - a_x)\cdot m_x}$$
The function arguments will be `mx` and `ax`. Just assume that you have a vector of `ax` for now. Or go ahead and make yourself some test data, like so:

```{r}
mx <- morty |> 
  filter(sex == "female", year == 2000) |> 
  pull(mx)
ax <- mx_to_ax(mx,age=0:110, sex = rep("f",111))
ax
```

```{r}
mx_to_qx <- function(mx, ax){
  qx <- mx / (1 + (1 - ax) * mx)
  qx <- if_else(qx > 1, 1, qx)
  return(qx)
}

qx <- mx_to_qx(mx, ax)
qx
```

# Exercise:

Make an lx function, whose argument is qx, and that scales to an arbitrary radix (ergo two arguments: qx, and radix).

$$ \ell_{x+1} = \prod _0^{x} (1-q_x)\quad  \forall x > 0 $$
And the very first value can be 1. And the whole thing can be scaled by `radix` (if you want).

```{r}
qx_to_lx <- function(qx, radix = 1){
  lx <- cumprod(1 - qx)
  # lx <- c(1, lx[-length(lx)])
  lx <- dplyr::lag(lx, default = 1)
  lx <- radix * lx
  return(lx)
}
lx <- qx_to_lx(qx, radix = 1e5)
```

# next step

$$ d_x = \ell_x - \ell_{x+1}$$

$$ d_x = \ell_x q_x$$

```{r}
lxqx_to_dx <- function(lx, qx){
  dx <- lx * qx
  return(dx)
}
dx <- lxqx_to_dx(lx, qx)
dx
```

Our first guesses at $e_0$:

```{r}
sum(lx) / 1e5 - .5
age <- 0:110
sum(dx * (age + ax)) / sum(dx)
```

# derive Lx

easier to think about it way:
$$ L_x = \ell_{x+1} + a_xd_x$$

easier to program way:
$$ L_x = \ell_x - (1-a_x)d_x$$

$L_x$ can be thought of as lifetable exposure
```{r}
lxdx_to_Lx <- function(lx, dx, ax){
  Lx <- lx - (1 - ax) * dx
  return(Lx)
}
Lx <- lxdx_to_Lx(lx,dx,ax)
mx2 <- dx / Lx
sum(Lx)/lx[1]
sum(Lx[-1])/lx[2]
sum(Lx[-c(1:2)])/lx[3]
```
$$ T_x = \sum _{t=x} ^\omega L_t $$
```{r}
Lx_to_Tx <- function(Lx){
  Tx <- Lx |> 
    rev() |> 
    cumsum() |> 
    rev()
  return(Tx)
}
Tx <- Lx_to_Tx(Lx)
Tx / lx
```

# Challenge: calculate all the lifetables in `morty`

To do this, you will start with morty, 
and then filter out `sex == "total"` 
and then declare groups by sex and year
and then use our lifetable functions in order inside a single `mutate()` call
and then remove the groups.

Be brave, make some plots afterwards.

























