---
title: "session 2 notes"
author: "Tim Riffe"
date: "2026-07-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ready to go!

# filter down to just countries
```{r}
library(tidyverse)
library(countrycode) # install.packages("countrycode")
wpp <- read_csv("data/mxwpp.csv.gz")
wpp <-
  wpp |> 
  mutate(iso3 = countrycode(sourcevar = un_code,
                            origin = "un",
                            destination = "iso3c")) |> 
  filter(!is.na(iso3))

```

# Challenge: 
Calculate the lifetables using our functions in `02_lifetables.R` (the `lt_full()` function specifically), just like we did in the exercise solutions. The groups in the data are unique combinations of iso3, sex, and year. This might take a while to run, sorry.


```{r}
source("02_lifetables.R")
wpp_lt <-
  wpp |> 
  group_by(iso3, sex, year) |> 
  group_modify(~lt_full(.x, .y, radix = 1)) |> 
  ungroup()
```


# took timeout to talk about CDR as weighted mean

Approach 1: add up the deaths, and add up the exposures, then take the ratio
```{r}
ES <- read_csv("data/ES_mort.csv.gz")

ES |> 
  summarize(D = sum(deaths),
            E = sum(exposure), 
            .by = c(sex, year)) |> 
  mutate(CDR = D / E)
```

Approach 2:
Treat CDR as weighted mean

$$\bar{x} = \frac{\sum w_ix_i}{\sum w_i}$$ 

Identical to:
$$ p_i = \frac{w_i}{\sum w_i}$$
$$ \bar{x} = \sum p_i x_i$$
$$m_x = \frac{D_x}{E_x}$$

$$ \frac{\sum D_x}{\sum E_x} =  \frac{\sum m_xE_x}{\sum E_x}$$


```{r}
ES |> 
  mutate(
    px = exposure / sum(exposure, na.rm = TRUE), 
         .by = c(year, sex)) |> 
  summarize(CDR = 1000*sum(exposure * mx, na.rm = TRUE) / sum(exposure, na.rm = TRUE),
            CDR2 = 1000*sum(px * mx, na.rm = TRUE),
            .by = c(year, sex)) |> 
  filter(year %in% c(1908, 2023))
```

# Standardization:

```{r}
standards <- read_csv("data/standards_abridged.csv")
standards |> 
  ggplot(aes(x = age, y = pop, color = standard)) + 
  geom_step()
```

We have two problems using these standards with single age data. First, the standard comes in age groups, and second, the age range is different than our data. The standard thing to do is to treat the standard as a step function. That means that if a given age class is 5 years wide, that we can simply divide the value of the standard in that age class by 5 and assign the result to each of the single ages falling in that class. Programmatically, it might look like so:

```{r}
age_abr <- standards |> 
  pull(age) |> 
  unique()
nx = case_when(age_abr == 0 ~ 1,
               age_abr == 1 ~ 4,
               age_abr == 90 ~ 11,
               TRUE ~ 5)
age_table <- tibble(age_single = 0:100,
                    age = rep(age_abr, times = nx),
                    nx = rep(nx, times = nx))
eu_standard <-
  standards |> 
  right_join(age_table, 
             by = join_by(age),
             relationship = "many-to-many") |> 
  mutate(pop = pop / nx) |> 
  select(standard,
         age = age_single,
         pop) |> 
  filter(grepl(pattern = "European", x = standard)) |> 
  select(-standard)
  # filter(standard == "European (EU-27 plus EFTA 2011-2030) Std Million")
  # group_by(standard) |> 
  # summarize(check = sum(pop)) |> 
```

Challenge: age-standardize either the wpp or the Spanish mortality series. The steps are to start with the mortality dataset, then `left_join` the `eu_standard` (or a different one if you want) by `age`, then group by `sex`, `year` (and also `iso3` if using wpp), then calculate the age-standardized mortality rate inside `summarize()`, using the sum-product of `pop` and `mx` (recall `pop` sums to 1 already). Inside said `sum()`, you may want to include the argument `na.rm = TRUE` if using the Spanish dataset.

```{r}
CDRstd <- ES |> 
  filter(age <= 100) |> 
  # join european standard
  left_join(
    eu_standard,
    by = join_by(age)
  ) |> 
  # computing standardized cdr
  summarise(
    SDR = sum(mx * pop, na.rm = TRUE),
    .by = c(year, sex))
```

plot these:
```{r}
CDRstd |> 
  filter(sex != "total") |> 
  ggplot(aes(x = year, y = SDR, color = sex)) +
  geom_line()
```

I noted here the following relationship in lifetables:
The age standardized rate of the lifetable uses the survival curve ($L_x$, I guess) is equal to $1/e_0$. So, sooo, if you imagined up an exercise like: age standardize a mortality and then take the inverse (it's units can be thought of as years of life) then you you'll know whether the values themselves (of SDR) carry much meaning.

```{r}
CDRstd |> 
  filter(sex != "total") |> 
  ggplot(aes(x = 1 / SDR, fill = sex)) +
  geom_density(alpha = .5)
```

Here we see values that are way outside the plausible range, so all this tells us is that the actual units of the SDR don't transmit the same level of meaning as a fully calculated life table.

# Decomposition

If we want to compare two CDRs, then we can standardize the rates using the average structure (direct standardization) or we can standardize the structure using the average rates (indirect standardization). Or, instead of standardizing those quantities, we could standardize the **differences** in rates and the **differences** in structure. If we do this, the sum of those two standardized differences is equal to the gap in CDR.

Formally: the difference in CDR can be split into two pieces. 

A rate-difference component:
$$ \Delta^{rates} = \sum\frac{1}{2}(str^1 + str^2) \cdot (m_x^2 - m_x^1)$$
And a structure difference component:
$$ \Delta^{str} = \sum\frac{1}{2}(m_x^1 + m_x^2) \cdot (str^2 - str^1)$$
And then we have that:
$$ \Delta = CDR^2 - CDR^1 = \Delta^{str} + \Delta^{rates} $$
This is called Kitagawa decomposition. Many kinds of decomposition relate to this basic pattern.

```{r}
lt_sd <- function(dx, age, ax){
  dx       <- dx / sum(dx)
  age_mid  <- age + ax
  e0       <- sum(dx * age_mid)
  distance <- age_mid - e0
  lt_var   <- sum(distance ^ 2 * dx)
  sd_lt    <- sqrt(lt_var)
  return(sd_lt)
}
```
Test it:

```{r}
ESlt |> 
  summarize(sd = lt_sd(dx, age, ax),
            e0 = first(ex),
           .by = c(year, sex)) |> 
  ggplot(aes(x = e0, y = sd, color = sex)) +
  geom_point()
```

















