---
title: "Session 4 notes"
author: "Tim Riffe"
date: "2026-07-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/04_data_prep.R")
```

Discounting stationary population structure for stable growth. This is the default stationary age structure. Note, Lx is what we would get if we took a census in 110 years, adding in the same number of sims per year, and holding the mortality pattern fixed.

```{r}
Lx <- LT |> 
  filter(sex == "female") |> 
  select(age, Lx)
Lx |> 
ggplot(mapping = aes(x = age, y = Lx / sum(Lx))) +
  geom_line()
```

`cross_join()` does a greedy data expansion, matching all unique values. Kind of like `expand_grid()`

```{r}
library(colorspace)
# hcl_palettes(plot = TRUE)
r <- seq(-.03,.03,by = .005) 
r <- tibble(r)
r |> 
  cross_join(Lx) |> 
  mutate(Cx = Lx * exp(-r * (age+.5)),
         Cx = Cx / sum(Cx),
         .by = r) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cx, 
                       color = r,
                       group = r)) +
  geom_line() +
  scale_color_binned_diverging()
```

But now, let's see how to mechanically project out to outcomes sort of like this.

We will make a Leslie matrix to do this. It's not the only way to put together the pieces, but it's somehow easier to keep track of everything if we stick all the demographic parameters in the same matrix.

A Leslie matrix has fertility in its upper row and survival in its subdiagonal.
```{r}
Sx <-
  Lx |> 
  mutate(Lx_next = lead(Lx),
         Sx = Lx_next / Lx) |> 
  filter(!is.na(Sx)) |> 
  pull(Sx)

L <- diag(Sx)
Les <- rbind(0,L) |> cbind(0)
```

The survival part of the Leslie matrix is sort of easier to wrap you mind around. You just to manage to put the right notion of survival probabilities on a subdiagnoal.

The fertility top row of the matrix has different pieces in it.
1. we need to average fertility rates between consecutive ages, to match how one age class moves into the next one in a projection.
2. we need to discount for the survival of women.
3. the births themselves need a discount to make it to the end of the year.
```{r}

asfr <-
ES2014 |> 
  filter(sex == "female") |> 
  mutate(exposure = (pop1 + pop2) / 2) |> 
  select(age, exposure) |> 
  left_join(B |> select(-sex), by = join_by(age)) |> 
  mutate(births = if_else(is.na(births), 0, births),
         asfr = births / exposure)
```


















