---
title: "session 4 solutions"
author: "Tim Riffe"
date: "2026-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Challenge:
See if the path to stability is different, or how it's different if we specify a wider or tighter fertility distribution. In terms of total perturbation of the route or maybe deviation sizes or smoothness or the width of the wiggles, etc.

# Load data
```{r}
library(tidyverse)
library(expm)
source("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/04_data_prep.R")
Lx <- LT |> 
  filter(sex == "female") |> 
  select(age, Lx)
Sx <-
  Lx |> 
  mutate(Lx_next = lead(Lx),
         Sx = Lx_next / Lx) |> 
  filter(!is.na(Sx)) |> 
  pull(Sx)

L <- diag(Sx)
Les <- rbind(0,L) |> cbind(0)

SRB <- 1.05
PF <- 1 / (1 + SRB)
asfr <-
  ES2014 |> 
  filter(sex == "female") |> 
  mutate(exposure = (pop1 + pop2) / 2) |> 
  select(age, exposure) |> 
  left_join(B |> select(-sex), by = join_by(age)) |> 
  mutate(births = if_else(is.na(births), 0, births),
         asfr = PF * births / exposure,
         asfr2 = lead(asfr, default = 0),
         asfr = (asfr + asfr2) / 2) |> 
  pull(asfr)


L0      <- Lx |> pull(Lx) |> first()
survival_part <- L0 * (1 - ((1-c(Sx,0)) / 2))
```
Now let's make some fertility assumptions to test the convergence path idea:
```{r}
tfrf         <- sum(asfr)
asfr_narrow  <- tfrf * (asfr^2) / sum(asfr^2)
asfr_wide    <- tfrf * (sqrt(asfr)) / sum(sqrt(asfr))

# make top rows
Les1 <- Les2 <- Les3 <- Les
Les1[1, ] <- asfr * survival_part
Les2[1, ] <- asfr_narrow * survival_part
Les3[1, ] <- asfr_wide * survival_part
```
 
```{r}
cbind(asfr, asfr_narrow, asfr_wide) |> 
  matplot(type = 'l')
```
 
 
 Apply:
```{r}
horizon <- tibble(h = 0:500)
Pt <- 
  P |> 
  filter(sex == "female") |> 
  select(age, P0 = pop2) |> 
  cross_join(horizon) |> 
  arrange(h, age) |> 
  mutate(Pt1 = (Les1 %^% first(h)) %*% P0,
         Pt2 = (Les2 %^% first(h)) %*% P0,
         Pt3 = (Les3 %^% first(h)) %*% P0,
         .by = h)
```
 
Plot:
```{r}
Pt |> 
  select(-P0) |> 
  pivot_longer(Pt1:Pt3, names_to = "variant", values_to = "Pt") |> 
  filter(h == 85) |> 
  mutate(Cx = Pt / sum(Pt), 
         .by = c(h,variant)) |> 
  ggplot(aes(x = age, 
             y = Cx, 
             color = variant, 
             group = variant)) +
  geom_line() +
  theme_minimal()
```
Observe: 80 years out, the wider fertility distribution has come far closer to the stable shape than the narrow distribution. The "echoes" are dampened greatly.

```{r}
Pt |> 
  select(-P0) |> 
  pivot_longer(Pt1:Pt3, names_to = "variant", values_to = "Pt") |> 
  filter(h == 500) |> 
  mutate(Cx = Pt / sum(Pt), 
         .by = c(h,variant)) |> 
   ggplot(aes(x = age, 
             y = Cx, 
             color = variant, 
             group = variant)) +
  geom_line() +
  theme_minimal()
```
 
```{r}
Pt |> 
  select(-P0) |> 
  pivot_longer(Pt1:Pt3, names_to = "variant", values_to = "Pt") |> 
  summarize(Pt = sum(Pt), .by = c(variant, h)) |> 
  mutate(rt = log(lead(Pt)/Pt), .by = variant) |> 
  ggplot(aes(x=h,y=rt,color=variant)) +
  geom_line()
```
 
 