---
title: "Session 4 notes"
author: "Tim Riffe"
date: "2026-07-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/04_data_prep.R")
```

Discounting stationary population structure for stable growth. This is the default stationary age structure. Note, Lx is what we would get if we took a census in 110 years, adding in the same number of sims per year, and holding the mortality pattern fixed.

```{r}
Lx <- LT |> 
  filter(sex == "female") |> 
  select(age, Lx)
Lx |> 
ggplot(mapping = aes(x = age, y = Lx / sum(Lx))) +
  geom_line()
```

`cross_join()` does a greedy data expansion, matching all unique values. Kind of like `expand_grid()`

```{r}
library(colorspace)
# hcl_palettes(plot = TRUE)
r <- seq(-.03,.03,by = .005) 
r <- tibble(r)
r |> 
  cross_join(Lx) |> 
  mutate(Cx = Lx * exp(-r * (age+.5)),
         Cx = Cx / sum(Cx),
         .by = r) |> 
  ggplot(mapping = aes(x = age, 
                       y = Cx, 
                       color = r,
                       group = r)) +
  geom_line() +
  scale_color_binned_diverging()
```

But now, let's see how to mechanically project out to outcomes sort of like this.

We will make a Leslie matrix to do this. It's not the only way to put together the pieces, but it's somehow easier to keep track of everything if we stick all the demographic parameters in the same matrix.

A Leslie matrix has fertility in its upper row and survival in its subdiagonal.
```{r}
Sx <-
  Lx |> 
  mutate(Lx_next = lead(Lx),
         Sx = Lx_next / Lx) |> 
  filter(!is.na(Sx)) |> 
  pull(Sx)

L <- diag(Sx)
Les <- rbind(0,L) |> cbind(0)
```

The survival part of the Leslie matrix is sort of easier to wrap you mind around. You just to manage to put the right notion of survival probabilities on a subdiagnoal.

The fertility top row of the matrix has different pieces in it.
1. we need to average fertility rates between consecutive ages, to match how one age class moves into the next one in a projection.
2. we need to discount for the survival of women.
3. the births themselves need a discount to make it to the end of the year.
4. we need to split out girl births from the total by assuming some SRB
```{r}
SRB <- 1.05
PF <- 1 / (1 + SRB)
asfr <-
  ES2014 |> 
  filter(sex == "female") |> 
  mutate(exposure = (pop1 + pop2) / 2) |> 
  select(age, exposure) |> 
  left_join(B |> select(-sex), by = join_by(age)) |> 
  mutate(births = if_else(is.na(births), 0, births),
         asfr = PF * births / exposure,
         asfr2 = lead(asfr, default = 0),
         asfr = (asfr + asfr2) / 2) |> 
  pull(asfr)

L0 <- Lx |> pull(Lx) |> first()
                      # TR: decided to discount mother
                      # mort by 2, since it's possible
                      # to give birth and then die in 
                      # same year
top_row <- asfr * L0 * (1 - ((1-c(Sx,0)) / 2))
Les[1, ] <- top_row
```

We can extract the "intrinsic" growth rate $r$ from the Leslie matrix, which just exploits standard eigen analysis, which is a rather technical topic for a course like this, but just to point out, it's a predetermined property of the demographic parameters themselves. And this is now demonstrated because we were somehow able to extract it from the Leslie matrix all by itself.
```{r}
r <- eigen(Les)$values |> first() |> Re() |> log()
```

How to do the mechanical projection
```{r}
P0 <- 
  P |> 
  filter(sex == "female") |> 
  pull(pop2)
plot(0:110,P0,type='l')
lines(0:110,Les %*% P0,col="red")
lines(0:110,Les %*% (Les %*% P0),col="blue")

Les %*% (Les %*% P0) 
(Les %*% Les) %*% P0

library(expm) # install.packages("expm")
 (Les %^% 10) %*% P0
```

Now demonstrate the long horizon projection using tidy tricks:

```{r}
horizon <- tibble(h = 0:200)
Pt <- 
  P |> 
  filter(sex == "female") |> 
  select(age, P0 = pop2) |> 
  cross_join(horizon) |> 
  arrange(h, age) |> 
  mutate(Pt = (Les %^% first(h)) %*% P0,
         .by = h)
```

How does the population structure stabilize in 20-year time steps:

```{r}
Pt |> 
  filter(h %% 20 == 0) |> 
  mutate(Cx = Pt / sum(Pt),
         .by = h) |> 
  ggplot(aes(x = age, y = Cx, color = h, group = h)) +
  geom_line() +
  scale_color_binned_sequential(begin = .5) +
  geom_line(data = Pt |> 
  filter(h == 0) |> 
  mutate(Cx = Pt / sum(Pt)), color = "red") +
  geom_line(data = Lx |> mutate(Cx = Lx * exp(-r*(age+.5)),
                                Cx = Cx / sum(Cx),
                                h=200), color = "black", linewidth = 2)
```

```{r}
Pt |> 
  summarize(Pt = sum(Pt), .by = h) |> 
  mutate(CGR = log(dplyr::lead(Pt)/Pt)) |> 
  ggplot(aes(x = h, y = CGR)) +
  geom_line() +
  geom_hline(yintercept = r, color = "red")
```

# Challenge:
See if the path to stability is different, or how it's different if we specify a wider or tighter fertility distribution. In terms of total perturbation of the route or maybe deviation sizes or smoothness or the width of the wiggles, etc.





