---
title: "Session 5 notes"
author: "Tim Riffe"
date: "2026-07-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# First, load packages and our lifetable stuff

```{r}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/02_lifetables.R")
```

# read in the mortality data:

```{r}
mx <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ESmx.csv.gz")

start_year <- 1970
```

# look at the surface:

```{r}
mx |> 
  ggplot(mapping = aes(x = year, y = age, z = log(mx))) +
  geom_contour_filled()
```

# plot alpha on top the log mortality rates
```{r}
alpha <-
  mx |> 
  filter(year >= start_year) |> 
  summarize(alphax = mean(log(mx)),
            .by = age)
mx |> 
  ggplot(aes(x = age, y = mx)) +
  geom_line(mapping = aes(color = year, group = year)) +
  scale_y_log10() +
  geom_line(data = alpha, mapping = aes(x=age,y=exp(alphax)), 
            color = "red", linewidth =2)
```

# get the matrix of centered log mortality

```{r}
clmx <- 
  mx |> 
  filter(year >= start_year) |> 
  left_join(alpha, by = join_by(age)) |> 
  mutate(lmx = log(mx),
         clmx = lmx - alphax)
```
Examine the pattern

```{r}
clmx |> 
  ggplot(aes(x = age, y = clmx, color = year, group = year)) +
  geom_line()
```

Turn CLMX into a matrix following the Lexis format. We spread years over the columns, then make the age column into row names and finally coerce to be a matrix:

```{r}
M <-
  clmx |> 
  select(year, age, clmx) |> 
  pivot_wider(names_from = year, values_from = clmx) |> 
  column_to_rownames("age") |> 
  as.matrix()
```

Now we can SVD this matrix:

```{r}
svdM <- svd(M)
d <- svdM$d
U <- svdM$u
V <- svdM$v

# we can exactly recreate the original matrix like so:
(U %*% diag(d) %*% t(V)) - M
```

Lets make the remaining two elements:
beta and kappa
```{r}
betax <- U[,1] / sum(U[,1])
kappat <- V[,1] * d[1] * sum(U[,1])
```

Get drift parameters and do the line projection 
```{r}
years <- mx |> pull(year) |> unique() |> sort()
drift <- (kappat[length(kappat)] - kappat[1]) / ((2014 - start_year) - 1)

horizon <- 1:30
kappat_hat <- horizon * drift + kappat[length(kappat)]
kappa_all <- c(kappat, kappat_hat)
plot(kappa_all, type = 'l')
```

Build tibbles and put the pieces together

```{r}
kappa <-
  tibble(year =start_year:2044,
         kappat = kappa_all)

beta <- tibble(age = 0:100,betax)

lc <- cross_join(kappa, beta) |> 
  left_join(alpha, by = join_by(age)) |> 
  mutate(mx = exp(alphax + betax * kappat))
```
Visualize the predictions:

```{r}
lc |> 
  ggplot(aes(x = year, y = age, z = log(mx))) +
  geom_contour_filled()

lc |> 
  filter(year > 2014) |> 
  ggplot(aes(x = age, y = mx, color = year, group = year)) +
  geom_line() +
  scale_y_log10()
```

Calculate lifetables for the LC prediction (the smooth kappa one)
Make a plot of life expectancy at birth.
```{r}
LTlc <- lc |> 
  mutate(sex = "female") |>
  group_by(year, sex) |>
  group_modify(~lt_full(.x,.y,radix = 1)) |> 
  ungroup()
LTlc |> 
  filter(age == 0) |> 
  ggplot() +
  aes(x = year, y = ex) +
  geom_line()


```
Contrast w period $e_0$ 1950-2014
```{r}
mx |> 
  mutate(sex = "female") |>
  group_by(year, sex) |>
  group_modify(~lt_full(.x,.y,radix = 1)) |> 
  filter(age == 0) |> 
   ggplot() +
  aes(x = year, y = ex) +
  geom_point(color = "red") + 
  geom_line(data = LTlc |> filter(age == 0) )
```

Here are some packages with out-of-the-box mortality forecasting options:

`demography`, `forecast`, `StMoMo`.

Books:
Data Visualization:
<https://kieranhealy.org/publications/dataviz/>

Intro Demography text:
Preston book:
<https://www.wiley.com/en-us/shop/general-introductory-history/demography-measuring-and-modeling-population-processes-p-9781557864512>

Pressat book:
<https://archive.org/details/demographicanaly0000pres_l0b4>

Ken Wachter:
<https://www.hup.harvard.edu/books/9780674045576>

Keyfitz applied mathematical demography
<https://link.springer.com/book/10.1007/978-1-4757-1879-9>

Online resources: (r packages for methods, reproducibility packs, workshop materials)
<https://github.com/OpenScienceDemography>



