---
title: "Session 5 notes"
author: "Tim Riffe"
date: "2026-07-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# First, load packages and our lifetable stuff

```{r}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/02_lifetables.R")
```

# read in the mortality data:

```{r}
mx <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2026Module2/refs/heads/master/data/ESmx.csv.gz")
```

# look at the surface:

```{r}
mx |> 
  ggplot(mapping = aes(x = year, y = age, z = log(mx))) +
  geom_contour_filled()
```

# plot alpha on top the log mortality rates
```{r}
alpha <-
  mx |> 
  summarize(alphax = mean(log(mx)),
            .by = age)
mx |> 
  ggplot(aes(x = age, y = mx)) +
  geom_line(mapping = aes(color = year, group = year)) +
  scale_y_log10() +
  geom_line(data = alpha, mapping = aes(x=age,y=exp(alphax)), 
            color = "red", linewidth =2)
```

# get the matrix of centered log mortality

```{r}
clmx <- 
  mx |> 
  left_join(alpha, by = join_by(age)) |> 
  mutate(lmx = log(mx),
         clmx = lmx - alphax)
```
Examine the pattern

```{r}
clmx |> 
  ggplot(aes(x = age, y = clmx, color = year, group = year)) +
  geom_line()
```

Turn CLMX into a matrix following the Lexis format. We spread years over the columns, then make the age column into row names and finally coerce to be a matrix:

```{r}
M <-
  clmx |> 
  select(year, age, clmx) |> 
  pivot_wider(names_from = year, values_from = clmx) |> 
  column_to_rownames("age") |> 
  as.matrix()
```

Now we can SVD this matrix:

```{r}
svdM <- svd(M)
d <- svdM$d
U <- svdM$u
V <- svdM$v

# we can exactly recreate the original matrix like so:
(U %*% diag(d) %*% t(V)) - M
```

Lets make the remaining two elements:
beta and kappa
```{r}
betax <- U[,1] / sum(U[,1])
kappat <- V[,1] * d[1] * sum(U[,1])
```

Get drift parameters and do the line projection 
```{r}
years <- mx |> pull(year) |> unique() |> sort()
drift <- (kappat[length(kappat)] - kappat[1]) / ((2014 - 1950) - 1)

horizon <- -64:30
kappat_hat <- horizon * drift + kappat[length(kappat)]

plot(horizon, kappat_hat, type = 'l')
points(-64:0, kappat)
```





